Thesis: draft

### 2021-05-22 draft.tex
Update draft.tex
changed results: stubborn actors

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_10x100_01-0001_003-0008_50-50_PR_REC.png}
    \label{sbm_50-50}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10x100_10-001-0001_50-50_PR_REC.png}
    \label{sbm-ws_50-50}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) over time for balanced starting conditions.]{Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) over time for balanced starting conditions. \textbf{(a)} SBM; \textbf{(b)} SBM-WS. Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_50_50}
\end{figure}
\newpage
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_50_50} shows the distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$. The region on the $x$-axis close to zero describes neighborhoods in which none, or few actors, adopt opinion B (these are thus neighborhoods in which most actors adopt opinion A). Conversely, the region on the $x$-axis close to one describes neighborhoods in which a majority of the actors adopt opinion B. The distribution is obtained by imposing ten equispaced bins, $\Delta x = 0.1$ \cite{Perra2019}.

@@

\begin{figure}[H]
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_ER_001_50-50_PR_REC_20-80_8x7.png}
      \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10-006_PR_REC_20-80_8x7.png}
    \label{ws_20-80_op}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}    
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_01-0001_003-0008_PR_REC_20-80_8x7.png}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10-001-0001_PR_REC_20-80_8x7.png}
    \label{sbm-ws_20-80_op}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) over time for unbalanced starting conditions.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) over time for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_20_80}
\end{figure}
\noindent
Fig. \ref{ev_op_20_80} shows that the minority opinion A eventually vanishes. This is the case for the low clustering networks, with $P_\text{A}(500) = 0.001 \pm 0.003$ for the ER network, $P_\text{A}(500) = 0.002 \pm 0.003$ for the SBM with a low modularity and $P_\text{A}(500) = 0.003 \pm 0.006$ for the SBM with a high modularity. The high clustering in the WS and SBM-WS networks seems to slightly hamper this vanishing of the minority opinion with an average value of $P_\text{A}(500) = 0.007 \pm 0.014$ for the SBM-WS network (Fig. \ref{sbm-ws_20-80_op}) and an average value of $P_\text{A}(500) = 0.009 \pm 0.015$ for the WS network (Fig. \ref{ws_20-80_op}).\\ 
\newline
This result differs from results in \cite{Perra2019}, where the minority opinion was able to survive over time for all the network models, even for the PR filtering algorithm, and where a minority opinion prevalence at $t=500$ between $P_\text{A}(500) = 0.06$ and $P_\text{A}(500) = 0.09$ was reported \cite{Perra2019}. The difference in the results is most likely due to an interplay between the network size $N$, the average degree $\left<k\right>$ and the maximum length of the actual time-line $S$. The results that where obtained in \cite{Perra2019} are for networks with size $N = 10^5$ nodes and average degree $\left<k\right> \sim 6$. In this thesis results are obtained for $N=10^3$ and average degree $\left<k\right> \sim 10$. In both cases the maximum length of the actual time-line is $S=20$. Fig. \ref{ev_op_WS_k=6} shows the results for the evolution of the prevalence of opinion A for an initial $20/80$ opinion distribution and using the PR filtering algorithm; Fig. \ref{ws1000} displays results for a WS model with $N=10^3$ and $\left<k\right> = 6$; Fig. \ref{ws10000} shows results for a WS model with $N=10^4$ and $\left<k\right> = 6$. We see that the minority opinion A reaches higher final values of $P_0(500)$ compared to the case depicted in Fig. \ref{ws_20-80_op}, with a value of $P_\text{A}(500) = 0.027 \pm 0.023$ in Fig. \ref{ws1000} and a value of $P_\text{A}(500) = 0.030 \pm 0.018$ in Fig. \ref{ws10000}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_1000-6_PR_20-80_8x7.png}
    \label{ws1000}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10000-6_PR_20-80_8x7.png}
    \label{ws10000}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) over time for unbalanced starting conditions. A WS model with $N=10^3$ or $N=10^4$, $\left<k\right>=6$ and $\beta = 0.01$ is implemented. The PR filtering algorithm is used.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) over time for unbalanced starting conditions. The Watts-Strogatz network with $\left<k\right> = 6$ and $\beta = 0.01$ is depicted with \textbf{(a)} $N=10^3$; \textbf{(b)} $N=10^4$. In both networks the average clustering coefficient is $\left<cc\right> = 0.584 \pm 0.003$. The PR filtering filtering algorithm is used. The dotted black represents the initial prevalence of opinion A. The dotted grey line represents a vanished prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_WS_k=6}
\end{figure}
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_20_80} shows the distribution of the fraction of friends of actor $i$ with the same opinion B at $t=500$.  The REC and PR filtering algorithms are compared for the ER, WS, SBM and SBM-WS.\\
\newline
The results for the ER model and the SBM (both the high and low modularity configuration) look qualitatively the same. Similarly, the results for the WS model and the SBM-WS agree with each other as well. This difference in results between ER/SBM (negligible clustering) and WS/SBM-WS (high clustering) suggests that it is the clustering that drives these differences.

@@

We study the effect of a stubbornness parameter $r_i$ on two configurations: (i) fraction of actors completely stubborn, (ii) all actors partly stubborn. In both cases we study both the evolution of the prevalence of the opinions and the formation of echo chambers. This is done for balanced and unbalanced starting conditions. \\
\newline
\textbf{Completely stubborn actors}\\
\newline
We make a fraction of the actors ($f_s$) completely stubborn ($r_i = 1$). We go from no stubborn actors, fracRes $=0$, to all the actors are stubborn, fracRes $=1$. \\
\newline
We will first discuss the formation of echo chambers for balanced starting conditions, after which the results for the opinion evolution will be shared. Then we will discuss the formation of echo chambers and the evolution of the prevalence of the opinions for unbalanced starting conditions.
	\includegraphics[width=10cm, height=8cm]{echo_chamber_vs_stubbornness_SBM_frac_compl_stubb_PR_10x8.png}
	\caption[Formation of echo chambers versus fraction of completely stubborn actors for the PR filtering algorithm and balanced starting conditions.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for balanced starting conditions and using the PR filtering algorithm. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$.}
\label{echo_vs_frac_complStubb}
Fig. \ref{echo_vs_frac_complStubb} shows the formation of echo chambers in the system versus the fraction of completely stubborn actors, fracRes, when using the PR filtering algorithm for balanced starting conditions. The $y$-axis represents the formation of echo chambers in the same way as was explained for Fig. \ref{echo_vs_mod} in Subsection \ref{echoVSmod}. The dotted black line represents an echo chamber size equal to one. This means that the number of echo chambers remains constant over time.\\
The Erd\H{o}s-R\'{e}nyi model (ER) shows no appearance of echo chambers. Introducing a fraction of completely stubborn actors does not change this behavior. The stochastic block model with a low modularity behaves similar as the Erd\H{o}s-R\'{e}nyi model (no formation of echo chambers). The data points for the SBM with a low modularity are not visible in Fig. \ref{echo_vs_frac_complStubb}, since they lie behind the ER data points.\\
The other three networks (WS, SBM-WS and SBM with a high modularity) all show qualitatively the same behavior. If an increasing fraction of completely stubborn actors is introduced, the formation of echo chambers decreases with this increasing fraction of stubborn actors.\\
The WS and SBM-WS networks are the high clustering networks. For a small fraction of stubborn actors they show the largest echo chambers. The stochastic block model with a high modularity has smaller echo chambers than the high clustering networks, but still has some formation of echo chambers for fracRes $\leqslant 0.1$. This leads us to make the same conclusion as before: a community structure enhances the formation of echo chambers compared to a random network, however not as significant as networks with a high clustering coefficient.

@@

  \caption[Formation of echo chambers versus fraction of completely stubborn actors for the REF and REC filtering algorithms and an initial $50/50$ opinion distribution.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for an initial $50/50$ opinion distribution. \textbf{(a)} REF filtering algorithm; \textbf{(b)} REC filtering algorithm. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. This value represents the formation of echo chambers in the system (denoted with echo chamber size in the figure). Results are shown for four different network models, each with $N=10^3$ and average degree $\left<k\right> \sim 10$: WS with $\left<k\right> =10,\ \beta = 0.06$ and $\left<cc \right> = 0.557 \pm 0.007$; SBM-WS with 10 communities of size 100, each community has the following WS parameters: $\left<k\right> = 10,\ \beta = 0.01$ and $p_{\text{add}} = 0.001$, $\left<cc\right> = 0.553 \pm 0.004$ and $Q = 0.909 \pm 0.004$; SBM with 10 communities of size 100, a high modularity case ($p_{\text{cl}} = 0.1,\ p_{\text{add}} = 0.001$, $Q = 0.908 \pm 0.005$) and a low modularity case ($p_{\text{cl}} = 0.03,\ p_{\text{add}} = 0.008$, $Q = 0.220 \pm 0.007$) are depicted; ER with $p= 0.01$. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$.}

@@

Fig. \ref{echo_vs_frac_complStubb_REF-REC} shows the formation of echo chambers in the system versus the fraction of completely stubborn actors, fracRes, when using the REF and REC filtering algorithms. The results are obtained for an initial $50/50$ opinion distribution. The results for the REF and REC filtering algorithms behave qualitatively the same, but are different from those obtained with the PR filtering algorithm (Fig. \ref{echo_vs_frac_complStubb}). Instead of a decrease in the formation of echo chambers with an increasing fraction of completely stubborn actors, we first observe an increase in the formation of echo chambers, followed by a decrease. This is especially clear for the high clustering networks (WS and SBM-WS). The SBM with a high modularity displays a small bump. The ER and the SBM with a low modularity (data points for this model lie behind the ER data points) do not seem to have a significant bump.\\

@@

These results suggest that the interplay between the REF or REC filtering algorithm and the presence of a (small) fraction of stubborn actors enhances the formation of echo chambers (mainly for networks with a high clustering and/or a clear community structure). If fracRes $\geqslant 0.25$, the formation of echo chambers starts to decrease. This is probably due to the fact that too many actors become stubborn, so there are more and more actors that will not change opinions. This in turn increases the chance of having stubborn neighbors, each with different opinions, who cannot change opinions, which disables the formation of echo chambers.\\

@@

This discrepancy between the results for the PR filtering algorithm and those for the REF and REC filtering algorithms in combination with a small fraction of stubborn actors can be potentially explained with the help of the following simple example. Assume we have three actors that form a triangle. Two actors initially carry opinion A and one actor carries opinion B. Lets first discuss the PR filtering algorithm and assume that none of the actors are stubborn. In case of the PR filtering algorithm this triangle has a high chance of forming an echo chamber (which is obtained if all the actors adopt the same opinion). This can be seen as follows: the one actor with opinion B is likely to switch to opinion A, since all its friends carry this opinion and hence, this opinion will be strongly pushed on its time-line (the fact that the PR filtering algorithm wants to promote posts with opinion B does not matter here, since there are no neighbors with opinion B). The two actors with opinion A will have more chance of keeping that opinion than switching to opinion B in case of the PR filtering algorithm. This is because the PR filtering algorithm will promote the posts of the friend with the same opinion A over the posts of the friend with opinion B. The end result will be that there is a high chance that all the three actors will adopt opinion A and hence, form an echo chamber. Now assume that we make the actor with opinion B stubborn. This actor is no longer capable of switching to opinion A. However, the two actors with opinion A still have more chance of keeping this opinion in case of the PR filtering algorithm, because this filtering algorithm keeps on promoting the posts of the neighbor with the same opinion A, and hence reduces the visibility of the posts of the stubborn friend with opinion B. The system will not form an echo chamber, since the stubborn actor with opinion B cannot switch to opinion A and the PR filtering algorithm reduces the chance of changing the opinion of the actors with opinion A to opinion B. The conclusion is that, in case of the PR filtering algorithm, making a fraction of the actors stubborn reduces the ability of the system to form echo chambers compared to the case where there are no stubborn actors. What happens if we use the REF or REC filtering algorithms? Again we have one actor with opinion B and two actors with opinion A in the triangle. Assume first that none of the actors are stubborn. For the PR filtering algorithm, we argued that in this case, the triangle has a high chance of evolving towards an echo chamber. For the REF and REC filtering algorithms this chance is however smaller, which can be seen as follows. The actor with opinion B still has a high chance of switching to opinion A, because all its friends (which are the two other actors in the triangle) carry this opinion. However, the two actors with opinion A will now not have a higher chance of keeping this opinion than to switch to opinion B, as was the case for the PR filtering algorithm. Instead, they have an equal chance of keeping their opinion as to change to the other opinion. This is because both the REF and REC filtering algorithms do not discriminate between the two different opinions of the friends of the actors with opinion A. Both opinion A and opinion B have an equal chance of being promoted on the time-line of the actors with opinion A. This reduced tendency of the actors with opinion A to not change their opinion (compared to using the PR filtering algorithm) hampers the ability of the triangle to form an echo chamber. This would explain why, in case of no stubborn actors, the REF and REC filtering algorithms show a lower formation of echo chambers compared to the PR filtering algorithm. Now, lets again assume that we make the actor with opinion B stubborn. In case of the PR filtering algorithm this reduced the formation of echo chambers in the triangle. For the REF and REC filtering algorithms, things are different. The stubborn actor with opinion B will not change opinion. The two actors with opinion A, on the other hand, will now have a higher chance of switching to opinion B (which was not the case when using the PR filtering algorithm). This can be seen as follows. The REF and REC filtering algorithms do not discriminate between the two different opinions of the two friends of the actors with opinion A. If the actor with opinion B is, however, stubborn it will keep on pushing that opinion on the time-lines of the two actors with opinion A. This will eventually lead to an increase of posts with opinion B, compared to posts with opinion A, on the time-lines of the two actors with opinion A. The result will then be that the actors with opinion A will have a higher chance of switching to opinion B, which enhances the formation of echo chambers in the triangle. This would explain the increase in the formation of echo chambers for a small fraction of stubborn actors when using the REF or REC filtering algorithms. If too many actors become stubborn, this increased ability of the system to form echo chambers, in case of using the REF or REC filtering algorithms, starts to decrease which can also be seen with the help of the triangle example. If the fraction of stubborn actors becomes too high, we might end up with the case where both the actor with opinion B and one of the actors with opinion A are stubborn. This strongly reduces the ability to form echo chambers in the triangle since we can no longer have a final situation in which all the three actors in the triangle carry the same opinion.\\

@@

Lets now study the results for unbalanced starting conditions ($20/80$). Opinion A is the minority opinion, which makes opinion B the majority opinion. First, we discuss the formation of echo chambers. Then, we investigate the results for the evolution of the prevalence of the opinions.\\

>>>>

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_10x100_01-0001_003-0008_50-50_PR_REC.png}
    \label{sbm_50-50}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10x100_10-001-0001_50-50_PR_REC.png}
    \label{sbm-ws_50-50}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions.]{Evolution of the prevalence of opinion A ($P_{\text{A}}(t)$) for balanced starting conditions. \textbf{(a)} SBM; \textbf{(b)} SBM-WS. Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_50_50}
\end{figure}
\newpage
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_50_50} shows the distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$. The region on the $x$-axis close to zero describes neighborhoods in which none, or few actors, adopt opinion B (these are thus neighborhoods in which most actors adopt opinion A). Conversely, the region on the $x$-axis close to one describes neighborhoods in which a majority of the actors adopt opinion B. The distribution is obtained by imposing ten equispaced bins, $\Delta x = 0.1$ \cite{Perra2019}.

@@

\begin{figure}[H]
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_ER_001_50-50_PR_REC_20-80_8x7.png}
      \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10-006_PR_REC_20-80_8x7.png}
    \label{ws_20-80_op}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}    
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_01-0001_003-0008_PR_REC_20-80_8x7.png}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10-001-0001_PR_REC_20-80_8x7.png}
    \label{sbm-ws_20-80_op}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_20_80}
\end{figure}
\noindent
Fig. \ref{ev_op_20_80} shows that the minority opinion A eventually vanishes. This is the case for the low clustering networks, with $P_\text{A}(500) = 0.001 \pm 0.003$ for the ER network, $P_\text{A}(500) = 0.002 \pm 0.003$ for the SBM with a low modularity and $P_\text{A}(500) = 0.003 \pm 0.006$ for the SBM with a high modularity. The high clustering in the WS and SBM-WS networks seems to slightly hamper this vanishing of the minority opinion with an average value of $P_\text{A}(500) = 0.007 \pm 0.014$ for the SBM-WS network (Fig. \ref{sbm-ws_20-80_op}) and an average value of $P_\text{A}(500) = 0.009 \pm 0.015$ for the WS network (Fig. \ref{ws_20-80_op}).\\ 
\newline
This result differs from results in \cite{Perra2019}, where the minority opinion was able to survive over time for all the network models, even for the PR filtering algorithm, and where a minority opinion prevalence at $t=500$ between $P_\text{A}(500) = 0.06$ and $P_\text{A}(500) = 0.09$ was reported \cite{Perra2019}. The difference in the results is most likely due to an interplay between the network size $N$, the average degree $\left<k\right>$ and the maximum length of the actual time-line $S$. The results that where obtained in \cite{Perra2019} are for networks with size $N = 10^5$ nodes and average degree $\left<k\right> \sim 6$. In this thesis results are obtained for $N=10^3$ and average degree $\left<k\right> \sim 10$. In both cases the maximum length of the actual time-line is $S=20$. Fig. \ref{ev_op_WS_k=6} shows the results for the evolution of the prevalence of opinion A for an initial $20/80$ opinion distribution and using the PR filtering algorithm; Fig. \ref{ws1000} displays results for a WS model with $N=10^3$ and $\left<k\right> = 6$; Fig. \ref{ws10000} shows results for a WS model with $N=10^4$ and $\left<k\right> = 6$. We see that the minority opinion A reaches higher final values of $P_0(500)$ compared to the case depicted in Fig. \ref{ws_20-80_op}, with a value of $P_\text{A}(500) = 0.027 \pm 0.023$ in Fig. \ref{ws1000} and a value of $P_\text{A}(500) = 0.030 \pm 0.018$ in Fig. \ref{ws10000}.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_1000-6_PR_20-80_8x7.png}
    \label{ws1000}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10000-6_PR_20-80_8x7.png}
    \label{ws10000}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. A WS model with $N=10^3$ or $N=10^4$, $\left<k\right>=6$ and $\beta = 0.01$ is implemented. The PR filtering algorithm is used.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. The Watts-Strogatz network with $\left<k\right> = 6$ and $\beta = 0.01$ is depicted with \textbf{(a)} $N=10^3$; \textbf{(b)} $N=10^4$. In both networks the average clustering coefficient is $\left<cc\right> = 0.584 \pm 0.003$. The PR filtering filtering algorithm is used. The dotted black represents the initial prevalence of opinion A. The dotted grey line represents a vanished prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_WS_k=6}
\end{figure}
\noindent
\textbf{Echo chambers}\\
\newline
Fig. \ref{echo_20_80} shows the distribution of the fraction of friends of actor $i$ with the same opinion B at $t=500$.  The REC and PR filtering algorithms are compared for the ER, WS, SBM and SBM-WS.\\
\newline
The results for the ER model and the SBM (both the high and low modularity configuration) look qualitatively the same. Similarly, the results for the WS model and the SBM-WS agree with each other as well. This difference in results between ER/SBM (negligible clustering) and WS/SBM-WS (high clustering) suggests that it is the clustering that drives these differences.

@@

We study the effect of a resistance parameter $r_i$ on two configurations: (i) fraction of actors completely stubborn, (ii) all actors partly stubborn. For both configurations, we study the evolution of the prevalence of the opinions and the formation of echo chambers. This is done for balanced and unbalanced starting conditions.

\subsubsection{Completely stubborn actors}\label{complStubbAct}

We make a fraction of the actors ($f_s$) completely stubborn ($r_i = 1$). We go from no stubborn actors, $f_s=0$, to all the actors are stubborn, $f_s=1$. \\
\newline
First, we discuss the formation of echo chambers for balanced starting conditions, after which the results for the opinion evolution are shared. Then, the formation of echo chambers and the evolution of the prevalence of the opinions for unbalanced starting conditions are discussed.\\
\newline
\textbf{Balanced starting conditions}\\
\newline
We start with a configuration in which $50 \%$ of the actors is given opinion A and $50 \%$ is given opinion B. Fig. \ref{echo_vs_frac_complStubb} shows the formation of echo chambers versus the fraction of stubborn actors, $f_s$. The PR filtering algorithm is used.

\begin{figure}[H]
	\includegraphics[width=10cm, height=8cm]{echo_chamber_vs_stubbornness_SBM_frac_compl_stubb_PR_10x8.png}
	\captionsetup{format=plain}
	\caption[Formation of echo chambers versus fraction of stubborn actors for the PR filtering algorithm and balanced starting conditions.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). The black dotted line represents a constant amount of echo chambers over time.  The PR filtering algorithm is used. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$.}
\label{echo_vs_frac_complStubb}
\end{figure}
\noindent
The ER model shows no appearance of echo chambers. Introducing a fraction of stubborn actors does not change this behavior. The SBM with low modularity behaves similar as the ER model. The data points for the SBM with low modularity lie behind the ER data points and are not visible in Fig. \ref{echo_vs_frac_complStubb}. The other networks (WS, SBM-WS and SBM with high modularity) all show qualitatively the same behavior. If an increasing fraction of stubborn actors is introduced, the formation of echo chambers decreases.\\
\newline
The WS and SBM-WS networks have high clustering. For a small fraction of stubborn actors they show the largest echo chambers. The SBM with high modularity has smaller echo chambers than the high clustering networks, but still has some formation of echo chambers for $f_s \leqslant 0.1$. These results confirm that community structure enhances the formation of echo chambers compared to random networks, however not as significant as networks with high clustering.

@@

  \caption[Formation of echo chambers versus fraction of stubborn actors for the REF and REC filtering algorithms and balanced starting conditions.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} REF filtering algorithm; \textbf{(b)} REC filtering algorithm. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). The black dotted line represents a constant amount of echo chambers over time. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$.}

@@

Results for the REF and REC filtering algorithms (Fig. \ref{echo_vs_frac_complStubb_REF-REC}) behave qualitatively the same, but are different from those obtained with the PR filtering algorithm (Fig. \ref{echo_vs_frac_complStubb}). Instead of a decrease in the formation of echo chambers with increasing fraction of stubborn actors, we first observe an increase, followed by a decrease. This is especially visible for the high clustering networks (WS and SBM-WS). The SBM with high modularity displays a small bump. The ER and the SBM with low modularity (data points for this model lie behind the ER data points) do not seem to have a significant dependency on the fraction of stubborn actors.\\

@@

These results suggest that the interplay between the REF or REC filtering algorithms and the presence of a (small) fraction of stubborn actors enhances the formation of echo chambers (mainly for networks with high clustering or community structure). If $f_s \geqslant 0.25$, the formation of echo chambers starts to decrease. This is due to the fact that too many actors become stubborn. This increases the chance of having stubborn neighbors, each with different opinions, who cannot change opinions. Hence, the formation of echo chambers is disabled.\\

@@

The discrepancy between the results for the PR filtering algorithm and those for the REF and REC filtering algorithms in combination with a small fraction of stubborn actors may be explained with the help of the following example. Assume we have three actors that form a triangle. Two actors initially carry opinion A and one actor carries opinion B. Lets first discuss the PR filtering algorithm and assume that none of the actors are stubborn. This triangle then has a high chance of forming an echo chamber (which is obtained if all the actors adopt the same opinion), which can be seen as follows: the actor with opinion B is likely to switch to opinion A, since all its friends carry this opinion and hence, this opinion will be recommended on its time-line (the fact that the PR filtering algorithm wants to promote posts with opinion B does not matter, since there are no neighbors with opinion B). The two actors with opinion A will have more chance of keeping that opinion than to switch to opinion B. This is because the PR filtering algorithm will promote posts of the friend with the same opinion A over posts of the friend with opinion B. There is thus a high chance that all three actors will adopt opinion A and form an echo chamber. Now assume that we make the actor with opinion B stubborn. This actor is no longer capable of switching to opinion A. The two actors with opinion A still have more chance of keeping this opinion: the PR filtering algorithm keeps on promoting posts of the neighbor with the same opinion A, and hence reduces the visibility of posts of the stubborn friend with opinion B. The system will be less likely to form an echo chamber: the stubborn actor with opinion B cannot switch to opinion A and the PR filtering algorithm reduces the chance of changing the opinion of the actors with opinion A to opinion B. The conclusion is that, in case of the PR filtering algorithm, making a fraction of the actors stubborn reduces the ability of the system to form echo chambers compared to the case where there are no stubborn actors. Lets now study what happens if we use the REF or REC filtering algorithms. Again we have one actor with opinion B and two actors with opinion A in the triangle. Assume first that none of the actors are stubborn. For the PR filtering algorithm, we argued that the triangle has a high chance of evolving towards an echo chamber. For the REF and REC filtering algorithms this chance is smaller. This can be understood as follows. The actor with opinion B has high chance of switching to opinion A, because all its friends carry this opinion. However, the two actors with opinion A will not have a higher chance of keeping this opinion than to switch to opinion B, as was the case for the PR filtering algorithm. Instead, they have an equal chance of keeping their opinion as to change to the other opinion. This is because the REF and REC filtering algorithms do not discriminate between the two different opinions of the friends of the actors with opinion A. Both opinion A and opinion B have an equal chance of being promoted on the time-line of the actors with opinion A. This reduced tendency of the actors with opinion A to not change their opinion (compared to the PR filtering algorithm) hampers the ability of the triangle to form an echo chamber. This explains why, in case of no stubborn actors, the REF and REC filtering algorithms show a lower formation of echo chambers compared to the PR filtering algorithm. Now, lets again assume that we make the actor with opinion B stubborn. In case of the PR filtering algorithm, this reduced the formation of echo chambers in the triangle. For the REF and REC filtering algorithms, things are different. The stubborn actor with opinion B will not change opinion. The two actors with opinion A, on the other hand, will have a higher chance of switching to opinion B (which was not the case for the PR filtering algorithm). This can be seen as follows. The REF and REC filtering algorithms do not discriminate between the two different opinions of the friends of the actors with opinion A. If the actor with opinion B is, however, stubborn it will keep on pushing that opinion on the time-lines of the actors with opinion A. This will lead to an increase of posts with opinion B, compared to posts with opinion A. The actors with opinion A will have a higher chance of switching to opinion B, which enhances the formation of echo chambers in the triangle. This explains the increase in the formation of echo chambers for a small fraction of stubborn actors when using the REF or REC filtering algorithms. If too many actors become stubborn, this increased ability of the system to form echo chambers starts to decrease which can also be seen with the help of the triangle example. If the fraction of stubborn actors becomes too high, we might end up with both the actor with opinion B and one of the actors with opinion A that are stubborn. This reduces the ability to form echo chambers since we can no longer have a final situation in which all the three actors in the triangle carry the same opinion.\\

@@

\textbf{Unbalanced starting conditions}\\
